Staged Abstract Interpreters

### 2019-09-09 paper/30_stageconc.tex
polish

To implement the staged concrete interpreter, we replay the steps from the
instantiation of the unstaged concrete interpreter in Section \ref{unstaged_conc}.
However, now we use the @Rep@ type to annotate the value domains, environments and
stores, and redefine the staged version of monads and primitive operations.

>>>>

To implement the staged concrete interpreter, we replay the steps from the
instantiation of the unstaged counterpart in Section \ref{unstaged_conc}.
However, now we use the @Rep@ type to annotate the value domains, environments and
stores, and redefine the staged version of monads and primitive operations.

### 2019-09-09 paper/40_absinterp.tex
polish

The abstract domains operated by abstract interpreters usually can be
formulated by complete lattices.
A complete lattice over set $S$ is a tuple $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\top$ is the top element, $\bot$ is the
bottom element, $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines a type
class of stage polymorphic lattices. Similar to @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to
annotate the binding-times, where @R[_]@ wraps the data type @S@ and @Boolean@
in these operations.

@@

In this section, we simply instantiate @SPLattice@ with the flat binding-time
type @R[T] = T@ (i.e., @NoRep@).
An example of lattices we used in the rest of the paper is the power set
abstract domain (shown in Figure \ref{fig:powerset}): the ordering is the subset
relation, the join is the union operation, and the meet is the intersect
operation. The bottom element of a power set is empty set, and we do not have a
representation of top element for power set.  Other lattices used in the paper, such as
products and maps, can be implemented similarly or by lifting the existing
lattices element-wise or point-wise. Non-relational numerical abstract domains
such intervals can also be implemented in a stage polymorphic way.

@@

In this section, we follow the \textit{abstracting definitional interpreters}
(ADI) approach \cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our concrete
interpreter to an abstract interpreter.

@@

We first widen the concrete values to @AbsValue@. There are two variants of
@AbsValue@: numbers are lifted to a singleton object @IntTop@, which represents
the set of all numbers; closures remain the same. Then the type @Value@ is
defined as a set of @AbsValue@.  The address space is constrained to be finite
to ensure the computability of analysis -- for a simple monovariant analysis,
we directly use variable names for addresses. After defining the @Value@ and
@Addr@ type, the types of environments and stores are automatically lifted to
abstract environments and abstract stores, i.e., @Map[Ident, Addr]@ and
@Map[Addr, Set[AbsValue]]@, respectively.  
\begin{lstlisting}
  trait AbstractComponents extends Semantics {
    sealed trait AbsValue
    case object IntTop extends AbsValue
    case class CloV(lam: Lam, env: Env) extends AbsValue
    type Value = Set[AbsValue]; case class Addr(x: String)
    ...
  }
\end{lstlisting}

It is important that the abstract stores now map addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
runtime values.  This can be justified by the approximation nature
and nondeterminism in the abstract interpreter.  First, when
analyzing a conditional expression, we may not have enough information to
decide which branch will be taken, thus a sound treatment is to explore the both
branches. Also, at some later point, we need to compute the join of two path.
Another nondeterminism comes from function applications, for instance, @f(a)@,
the possible callees ($\lambda$ terms) of @f@ is computed approximately.  As a
result, we may retrieve a set of closures when dereferencing the address of @f@
and we need to apply all of them.

@@

\paragraph{Monads for Abstract Interpretation}
Compared with the concrete interpreter that uses reader and state effects, the
abstract interpreter further introduces the nondeterminism effect and another
reader and state effect for the two caches. The nondeterminism effect is
represented by the @Set[M[_], A]@ monad transformer, where @M@ is the inner
monad type being transformed, and @A@ is the type of elements in the set. We use a
@ReaderT@ for one cache that is not changed during one fixed-point
iteration, and use a @StateT@ for another cache that will be constantly updated
during the analysis.  \citet{DBLP:journals/pacmpl/DaraisLNH17} discuss
different permutations of the monad stack for abstract interpretation. The
following @AnsM@ type shows the monad stack for abstract interpretation (we use
\hl{light gray} to highlight what have been changed from the monad stack of
concrete interpretation):
\begin{lstlisting}[escapechar=!]
trait AbstractSemantics extends AbstractComponents {
  type R[T] = T
  type AnsM[T] = ReaderT[StateT[!\hl{SetT}![!\hl{ReaderT}![!\hl{StateT}![IdM, !\hl{Cache}!, ?], !\hl{Cache}!, ?], ?], Store, ?], Env, T]
  ...
}
\end{lstlisting}

Similar to the concrete scenario, we sketch the @flatMap@ implementation of the
@SetT@ transformer and omit the rest. The field @run@ encapsulated by the monad
is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and @A@ is the element
type of the set. The @flatMap@ method takes a function that transforms each
element of type @A@ into a @SetT[M, B]@ value, and then folds
all the transformed values into a single monadic value of type @Set[M, B]@.
\begin{lstlisting}
  case class SetT[M[_]: Monad, A](run: M[Set[A]]) {
    def flatMap[B](f: A => SetT[M, B]): SetT[M, B] =
      SetT(Monad[M].flatMap(run) { (s: Set[A]) =>
        s.foldLeft(SetT.empty[M, B])((acc, a) => acc ++ f(a)).run
      }); ... }
\end{lstlisting}

Such monad stack scheme leads to a different type of grounded result. In the concrete
setting, we only have a reader monad (for environments) and a state monad (for
stores). The reader effect is not persistent to the final result. Together
with the value produced by the interpreter, the final result type is a pair of
@Value@ and @Store@. However, under abstract interpretation, we have a @SetT@
inside of the environment monad and store monad. The @Set@ type becomes the
container type of the pairs of values and stores, i.e., @Set[(Value, Store)]@.
Also, note that the reader and state monad for the caches inhibits internally
of the nondeterminism monad; therefore, the final result type is paired
with a cache type from the state monad.
\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

@@

The value representation of lambda terms is still a defunctionalized closure
@CloV@, but we lift it to a singleton set.
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = Set(CloV(λ, ρ))
\end{lstlisting}

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is that now the first argument @funs@ is a set of closures (remember
that type @Value@ is an alias of @Set[AbsValue]@), so we can not directly
destruct it to a syntactic @Lam@ by pattern matching. Instead, we use an
auxiliary function @lift_nd@ that takes a set and lifts the elements in the set
into the monad stack. Then we can straightforwardly implement the @ap_clo@ in
monadic style, where the closures come from the @SetT@ and thus the
nondeterminism can be naturally handled. The light gray code shows what is
added from the concrete
@ap_clo@.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(funs: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](funs)}!
    α <- alloc(x)
    _ <- set_store(α → funs)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

\paragraph{Caching and Fixpoint Iteration}

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination.
The caching algorithm is also called \textit{co-inductive} caching or
\textit{truncated depth-first evaluation} \cite{Rosendahl:AbsIntPL}. It has
been used in other abstract interpreters or fixed-point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use a @in@ cache and a @out@ cache during
the depth-first evaluation. The @in@ cache stores the result from the last
iteration, and @out@ cache is used for updating in the current iteration. In the next
iteration, the last @out@ cache will be served as @in@ cache, and an empty cache
is plugged-in to the @out@ slot. Once the @out@ cache does not contain anything
new than the @in@ cache, the fixed-point is reached.

In our monad stack, the @in@ and @out@ caches are modeled by reader effects and
state effects, respectively. We first define several methods to manipulate
the two caches through the monad stack (the implementations are elided).
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

>>>>

The abstract domains operated by abstract interpreters usually can be
formulated by complete lattices.
A complete lattice over set $S$ is a tuple $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\top$ is the top element, $\bot$ is the
bottom element, $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines a type
class of stage polymorphic lattices. Similar to the @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to the
trait for annotating the binding-times, where @R[_]@ wraps the data type
@S@ and @Boolean@ in these operations.

@@

In this section, we simply instantiate @SPLattice@ with the flat binding-time
type @R[T] = T@ (i.e., @NoRep@); the result is the trait @Lattice[S]@.
An example of the lattices we used in the rest of the paper is the power set
abstract domain (shown in Figure \ref{fig:powerset}). For two power sets,
they are ordered by the subset relation. We use set union to compute their
join, and set intersection to compute their meet.  The bottom element of a
power set is empty set, and we do not have a representation of top element for
power set.  Other lattices used in the paper, such as products and maps, can be
implemented similarly or by lifting the existing lattices element-wise or
point-wise.  Non-relational numerical abstract domains such intervals can also
be implemented in a stage polymorphic way.

@@

In this section, we follow the abstracting definitional interpreters
(ADI) approach \cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our monadic
concrete interpreter to an monadic abstract interpreter.

@@

We first widen the concrete value domain to @AbsValue@. There are two variants of
@AbsValue@: 1) numbers are lifted to a singleton object @IntTop@, which represents
the set of all numbers; 2) closures remain the same. Then, the type @Value@ is
redefined as a set of @AbsValue@ to account for approximation.  The address
space is constrained to be finite to ensure the computability of the analysis --
for a simple monovariant analysis, we directly use variable names for
addresses. After defining the @Value@ and @Addr@ type, the types of
environments and stores are automatically lifted to their abstract versions,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@, respectively.  
\begin{lstlisting}
  trait AbstractComponents extends Semantics {
    sealed trait AbsValue
    case object IntTop extends AbsValue
    case class CloV(lam: Lam, env: Env) extends AbsValue
    type Value = Set[AbsValue]; case class Addr(x: String)
    ... }
\end{lstlisting}

It is important that the abstract stores now map addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
runtime values.  This can be justified by the approximation nature and
nondeterminism during the analysis.  For example, when analyzing a
conditional expression, we may not have enough information to decide which
branch will be taken, thus a sound treatment is to explore the both branches.
Also, at some later point, we need to compute the join of two path.  Another
nondeterminism comes from function applications, for instance, @f(a)@, the
possible $\lambda$ terms of @f@ is computed approximately.  As a
result, we need to store all the possible callees, as well as we will
retrieve a set of closures when dereferencing an address.

@@

\paragraph{Monads for Abstract Interpretation}
Compared with the concrete interpreter that uses reader and state effects, the
abstract interpreter further introduces the nondeterminism effect and another
reader and state effect for the two caches. The nondeterminism effect is
represented by the @Set[M[_], A]@ monad transformer, where @M@ is the inner
monad type being transformed, and @A@ is the type of elements in the set. 
We use a @ReaderT@ for one cache that is not changed during one fixed-point
iteration, and use a @StateT@ for another cache that will be constantly updated
during the analysis.
\citet{DBLP:journals/pacmpl/DaraisLNH17} discuss different permutations of the
monad stack for abstract interpretation. The following @AnsM@ type shows the
monad stack (we use \hl{light gray} to highlight what have been changed from
the monad stack of concrete interpretation):
\begin{lstlisting}[escapechar=!]
trait AbstractSemantics extends AbstractComponents {
  type R[T] = T
  type AnsM[T] = ReaderT[StateT[!\hl{SetT}![!\hl{ReaderT}![!\hl{StateT}![IdM, !\hl{Cache}!, ?], !\hl{Cache}!, ?], ?], Store, ?], Env, T]
  ... }
\end{lstlisting}

Similar to the concrete scenario, we sketch the @flatMap@ implementation of the
@SetT@ transformer and omit the rest. The field @run@ encapsulated by the monad
is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and @A@ is the element
type of the set. We first apply @flatMap@ on the inner monad to obtain the
set @s@. Then, we use the function @f@ to transform every element of type @A@
into a monadic value of type @Set[M, B]@. Finally, we fold all the transformed
values into a single monadic value of type @Set[M, B]@ by appending all of them.
The initial value of the fold is an empty @SetT@.
\begin{lstlisting}
  case class SetT[M[_]: Monad, A](run: M[Set[A]]) {
    def flatMap[B](f: A => SetT[M, B]): SetT[M, B] =
      SetT(Monad[M].flatMap(run) { (s: Set[A]) =>
        s.foldLeft(SetT.empty[M, B])((acc, a) => acc ++ f(a)).run
      }); ... }
\end{lstlisting}

This monad stack scheme (@AnsM[T]@) leads to a different type of grounded
result. Recall that in the concrete setting, we only have a reader monad (for
environments) and a state monad (for stores). The reader effect is not
persistent to the final result. Hence, together with the value produced by the
interpreter, the final result type is just a pair of @Value@ and @Store@.
However, under the new monad stack for abstract interpretation, we have a
@SetT@ inside of the environment monad and store monad. Therefore, the type
@Set@ becomes the container type of the pairs of values and stores, i.e.,
@Set[(Value, Store)]@.  Also, note that the reader and state monad for the
caches both inhibit internally of the nondeterminism monad; as the result, the
final result type is paired the set of value-stores with the cache type from
that state monad, as shown below.
\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

@@

The value representation of lambda terms is still a defunctionalized closure of
type @CloV@, but we lift it to a singleton set of @AbsValue@ (remember that
type @Value@ is an alias of @Set[AbsValue]@).
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = Set(CloV(λ, ρ))
\end{lstlisting}

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is that now the first argument of @ap_clo@ is a set of closures
@funs@, so we can not directly destruct it to a syntactic @Lam@ by pattern
matching. Instead, we use an auxiliary function @lift_nd@ that takes a set and
lifts the elements in the set into the monad stack. Then we can
straightforwardly implement the @ap_clo@ in monadic style, where the closures
come from the monads and thus the nondeterminism can be naturally handled. The
light gray code shows what is added from the concrete @ap_clo@.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(funs: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](funs)}!
    α <- alloc(x)
    _ <- set_store(α → funs)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

\paragraph{Caching and Fixpoint Iteration}
As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination.
The caching algorithm is also called a \textit{co-inductive} caching or
\textit{truncated depth-first evaluation} \cite{Rosendahl:AbsIntPL}. It has
been used in other abstract interpreters or fixed-point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use a @in@ cache and a @out@ cache during
the depth-first evaluation. The @in@ cache stores the result from the last
iteration, and @out@ cache is used for updating in the current iteration. In
the next iteration, the last @out@ cache will be served as the @in@ cache, and
an empty cache is plugged-in to the @out@ slot. Once the @out@ cache does not
contain any new information compared with the @in@ cache, the fixed-point is
reached.
In our monad stack, the @in@ and @out@ caches are modeled by the reader monad and
state monad, respectively. We first define several methods to manipulate
the two caches through the monad stack (the implementations are elided).
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}
